---
title: "BC Student Grade-to-Grade Transitions"
output: html_notebook
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(scales)
```
## Introduction

It is critical for the effectivess of any educational program that its students 
progress smoothly between grade levels. The BC Ministry of Education has released
[data](https://catalogue.data.gov.bc.ca/dataset/bc-schools-grade-to-grade-transition) 
on grade-to-grade transitions categorized by a number of discriminating features, of 
which we will examine

  1. `BASE_YEAR` / `SCHOOL_YEAR`: the transition reference year
  2. `PUBLIC_OR_INDEPENDENT`: public school vs independent school identifiers
  3. `DISTRICT_NUMBER` / `DISTRICT_ NAME`: school district identifier
  4. `STUDENT_GRADE`: student's grade level in `BASE_YEAR`
  5. `SUB_POPULATION`: various methods of breaking down student populations (indigenous/non-indigenous, male/female, etc.)
  6. `DENOMINATOR_STUDIED_COHORT`: number of students in a particular cohort
  7. `NUMERATOR_NUM_SUCCESSFUL`: number of students in a particular cohort who successfully transitioned between grades

We first load the data for `BASE_YEAR` 2015 in order to perform preliminary analyses.

```{r}
df = read.csv('grade_transition.csv')
df2 = df[df$BASE_YEAR=='2015/2016',]
df3 = df2[(df2$DATA_LEVEL=='District Level' | 
            df2$PUBLIC_OR_INDEPENDENT=='BC Independent School' & df2$DATA_LEVEL=='Province Level')
          & (df2$SUB_POPULATION=='Indigenous' | df2$SUB_POPULATION=='Non Indigenous'),]

gtdf = df3[, c('BASE_YEAR', 'PUBLIC_OR_INDEPENDENT', 'DISTRICT_NUMBER',
                  'STUDENT_GRADE', 'SUB_POPULATION', 
                  'DENOMINATOR_STUDIED_COHORT', 'NUMERATOR_NUM_SUCCESSFUL')]

head(gtdf)
```

In order to fit models to the data, we process it into a more convenient format.

```{r}
gtdf$BASE_YEAR = as.numeric(substr(gtdf$BASE_YEAR, 1, 4))

gtdf$PUBLIC_OR_INDEPENDENT = as.character(gtdf$PUBLIC_OR_INDEPENDENT)
gtdf[gtdf$PUBLIC_OR_INDEPENDENT=='BC Independent School',]$PUBLIC_OR_INDEPENDENT = 0
gtdf[gtdf$PUBLIC_OR_INDEPENDENT=='BC Public School',]$PUBLIC_OR_INDEPENDENT = 1
gtdf$PUBLIC_OR_INDEPENDENT = as.factor(gtdf$PUBLIC_OR_INDEPENDENT)

gtdf$SUB_POPULATION = as.character(gtdf$SUB_POPULATION)
gtdf[gtdf$SUB_POPULATION=='Indigenous',]$SUB_POPULATION=1
gtdf[gtdf$SUB_POPULATION=='Non Indigenous',]$SUB_POPULATION=0

gtdf[is.na(gtdf$DISTRICT_NUMBER),]$DISTRICT_NUMBER=0
gtdf$DISTRICT_NUMBER = as.factor(gtdf$DISTRICT_NUMBER)

gtdf$DENOMINATOR_STUDIED_COHORT = as.character(gtdf$DENOMINATOR_STUDIED_COHORT)
gtdf$NUMERATOR_NUM_SUCCESSFUL = as.character(gtdf$NUMERATOR_NUM_SUCCESSFUL)

gtdf = gtdf[gtdf$DENOMINATOR_STUDIED_COHORT!='Msk' & gtdf$NUMERATOR_NUM_SUCCESSFUL!='Msk',]

gtdf$DENOMINATOR_STUDIED_COHORT = as.numeric(gtdf$DENOMINATOR_STUDIED_COHORT)
gtdf$NUMERATOR_NUM_SUCCESSFUL = as.numeric(gtdf$NUMERATOR_NUM_SUCCESSFUL)

colnames(gtdf) = c('base_year', 'is_public', 'district_number', 'student_grade',
                   'is_indigenous', 'total', 'successful')
rownames(gtdf) = 1:length(gtdf$base_year)

gtdf$is_public = as.numeric(gtdf$is_public)
gtdf$is_indigenous = as.numeric(gtdf$is_indigenous)
gtdf$success_rate = gtdf$successful / gtdf$total
head(gtdf)
```

## Preliminary Analysis
In the exploratory stage, will use visualizations to identify patterns in the data
that may be quantifiable with statistical models. 

### Transition Rates for Indigenous and Non-Indigenous Students
```{r}
agg_df = gtdf %>% group_by(is_indigenous) %>% 
  summarise(total=sum(total), successful=sum(successful))
agg_df$success_rate = agg_df$successful / agg_df$total
agg_df$is_indigenous = as.character(agg_df$is_indigenous)
agg_df[agg_df$is_indigenous==1,]$is_indigenous = 'Indigenous'
agg_df[agg_df$is_indigenous==0,]$is_indigenous = 'Non-Indigenous'

ggplot(aes(is_indigenous, success_rate), data=agg_df) + 
  geom_bar(stat='identity', fill='indianred3') +
  geom_text(aes(label=paste(as.character(round(success_rate*100, 1)), '%')), vjust=-0.5) +
  scale_y_continuous(limits=c(0.95, 1), 
                     oob=rescale_none, 
                     labels=scales::percent) +
  xlab('') + ylab('Transition Rate') +
  ggtitle('Transition Rates for Indigenous and Non-Indigenous Students')
```

It is clear that indigenous student face a considerably lower overall transition 
rate than non-indigenous students - this will need to be explored further.

```{r message=FALSE, warning=FALSE}
agg_df = gtdf %>% group_by(is_indigenous, student_grade) %>% 
  summarise(total=sum(total), successful=sum(successful))
agg_df$success_rate = agg_df$successful / agg_df$total
agg_df$is_indigenous = as.character(agg_df$is_indigenous)
agg_df[agg_df$is_indigenous==1,]$is_indigenous = 'Indigenous'
agg_df[agg_df$is_indigenous==0,]$is_indigenous = 'Non-Indigenous'

agg_df = rename(agg_df, 'Sub-Population'=is_indigenous)

ggplot(aes(student_grade, success_rate), data=agg_df) + 
  geom_line(aes(color=`Sub-Population`)) + 
  geom_point(aes(color=`Sub-Population`)) +
  geom_text(aes(label=paste(as.character(round(success_rate*100, 0)), '%'), color=`Sub-Population`), vjust=-0.5) +
  scale_y_continuous(labels=scales::percent) +
  xlab('Grade Level (Base)') + ylab('Transition Rate') +
  ggtitle('Transition Rates by Grade for Indigenous and Non-Indigenous Students')
```
While the discrepancy between the groups exists to some extent at all grade levels, 
it is most dramatic in the final 3 years, likely suggesting that many of these 
non-transitioning students are, in fact, dropping out of school. This is a hypothesis
that ought to be explored in its on right.

### Transition Rates by School Type
```{r}
agg_df = gtdf %>% group_by(is_public) %>% 
  summarise(total=sum(total), successful=sum(successful))
agg_df$success_rate = agg_df$successful / agg_df$total
agg_df$is_public = as.character(agg_df$is_public)

agg_df[agg_df$is_public==2,]$is_public = 'Public'
agg_df[agg_df$is_public==1,]$is_public = 'Independent'

ggplot(aes(is_public, success_rate), data=agg_df) + 
  geom_bar(stat='identity', fill='indianred3') + 
  geom_text(aes(label=paste(as.character(round(success_rate*100, 1)), '%')), vjust=-0.5) +
  scale_y_continuous(limits=c(0.95, 1), 
                     oob=rescale_none, 
                     labels=scales::percent) +
  xlab('') + ylab('Transition Rate') +
  ggtitle('Transition Rates by School Type')
```
The overall difference between public and independent schools appears to be minimal.
```{r message=FALSE, warning=FALSE}
agg_df = gtdf %>% group_by(is_public, student_grade) %>% 
  summarise(total=sum(total), successful=sum(successful))
agg_df$success_rate = agg_df$successful / agg_df$total
agg_df$is_public = as.character(agg_df$is_public)
agg_df[agg_df$is_public==2,]$is_public = 'Public'
agg_df[agg_df$is_public==1,]$is_public = 'Independent'

agg_df = rename(agg_df, 'School Type'=is_public)

ggplot(aes(student_grade, success_rate), data=agg_df) + 
  geom_line(aes(color=`School Type`)) + 
  geom_point(aes(color=`School Type`)) +
  geom_text(aes(label=paste(as.character(round(success_rate*100, 0)), '%'), color=`School Type`), vjust=-0.5) +
  scale_y_continuous(labels=scales::percent) +
  xlab('Grade Level (Base)') + ylab('Transition Rate') +
  ggtitle('Transition Rates by Grade by School Type')
```
Again, we observe a drop in the final year, although not nearly as dramatic as
with indigenous students (92% overall public school vs 82% overall indigenous). 
It is possible that students in independent schools are more likely to see 12th 
grade as a necessary prerequisite for attending university, while students in 
public schools are more likely to forgo their final year of school in order to
join the workforce - this remains to be further investigated.

## Statistical Analysis
### Baseline Logistic Regression
We will establish a baseline for statistical analysis with a logistic regression
on `is_public` (binary), `is_indigenous` (binary) and `student_grade` (treated
as an ordinal discrete predictor)

```{r}
mod1 = glm(cbind(successful, total-successful) ~ is_public + student_grade + is_indigenous, 
           family='binomial',
           data=gtdf)

summary(mod1)
```

All parameters are statistically significant at the 95% confidence level. We interpret
the fitted coefficients in terms of their multiplicative effects on the 
odds $\frac{p}{1-p}$ of transitioning between grades.
```{r}
exp(mod1$coefficients)-1
```
  1. Attending public school increases the odds of transitioning by 6%
  2. Each increase in grade level decreases the odds of transitioning by 14%
  3. Being indigenous decreases the odds of transitioning by 50%
  
Because the tail probability for `is_public` is relatively large compared to 
`student_grade` and `is_indigenous` (and drawing on our previous intuition), we 
will attempt to remove that parameter from the model.

```{r}
mod2 = glm(cbind(successful, total-successful) ~ student_grade + is_indigenous, 
           family='binomial',
           data=gtdf)
summary(mod2)
```
The simplified model $\mathcal{M}_2$is nested within the previous model $\mathcal{M}_1$
, and it therefore has a `\chi^2_1` distribution under the null hypothesis that 
$\mathcal{M}_2$ is not an adequate simplification of $\mathcal{M}_1$

```{r}
pchisq(mod2$deviance - mod1$deviance,
       mod2$df.residual - mod1$df.residual,
       lower.tail=FALSE)
```
Thus, we would reject the inadequacy of the simplification at the 95% confidence
level, but fail to reject at 99%. As we are considering baselines, we shall merely
make note of that at this stage.

### Variable Interactions
